---
title: "Class 5: Data Viz with ggplot"
author: "Jennifer Thai (PID: A17893762)"
format: pdf
---

Today we are exploring the **ggplot** package and how to make nice figures in R.

There are lots of ways to make figures and plot in R. These include:

- so called "base" R
- and add on packages like **ggplot2**

Here is a simple "base" R plot.

```{r}
head(cars)
```

We can simply pass to the `plot()` function.

```{r}
plot(cars)
```

> Key-point: Base R is quick but not so nice looking in some folks eyes

Let's see how we can plot this with **glotplot2**...

1st I need to install this add-on package. For this we use the `install.packages()` function - **WE DO THIS IN THE CONSOLE, NOT our report**. This is a one time only deal.

2nd We need to load the package with the `library()` function every time we want to use it.


```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is composed of at least 3 layers:

- **data** (i.e. a data.frame with the things you want to plot)
- aesthetics **aes()** that map the columns of data to your plot features (i.e. aesthetics)
- geoms like **geom_point()** that srt how the plot appears

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

> Key point: For simple "canned" graphs, base R is quicker, but as things get more custom and elaborate, then ggplot wins out...

Let's add more layers to our ggplot

Add a line showing the relationship between x and y
Add a title
Add custom axis labels "Speed (MPH)" and "Distance(ft)"
Change the theme...

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Silly plot of Speed vs Stopping distance", 
       x = "Speed (MPH)", 
       y = "Distance (ft)") +
  theme_bw()
```

## Going further

Read some gene expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)

head(genes)
```

> Q1. How many genes are in this wee dataset?

```{r}
nrow(genes)
```
```{r}
ncol(genes)
```

> Q2. How many "up" regulated genes are there?

```{r}
sum( genes$State == "up" )
```

A useful function for counting up occurrences of things in a vector is the `table()` function.


```{r}
table( genes$State )
```

Make a v1 figure

```{r}
p <- ggplot(genes) +
  aes(x=Condition1, 
      y=Condition2,
      col=State) +
  geom_point() 
  
p
```

```{r}
p + scale_colour_manual(values =c("blue", "gray", "red")) +
  labs(title="Expression changes upon drug treatment",
       x= "Control (no drug)", 
       y= "Treatment (with drug)") +
  theme_bw()
```


## More Plotting 

Read in the gapminder dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

Let's have a wee peak

```{r}
head(gapminder, 3)
```

```{r}
tail(gapminder, 3)
```

> Q4. How many different country values are in this dataset?

```{r}
length( table(gapminder$country) )
```
> Q5. How many different continent values are in this dataset?

```{r}
length( unique(gapminder$continent) )
```
```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col= continent) +
  geom_point()
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col= continent, label= country) +
  geom_point() +
  geom_text()
```

I can use the **ggrepl** package to make more sensible labels here.
```{r}
library(ggrepel)

ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col= continent, label= country) +
  geom_point() +
  geom_text_repel()
```

I want a separate panel per continent

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col= continent, label= country) +
  geom_point() +
  facet_wrap(~continent)
```


ggplot2 offers several advantages over base R plots:

The main advantages of ggplot over base R plot are:

1. **Layered grammar**: ggplot2 builds plots by adding layers (data, aesthetics, geoms, themes), making complex plots easier to construct and modify step-by-step [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk).
2. **Consistency**: The same syntax and logic apply across different plot types, reducing the need to learn new functions for each visualization [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk).
3. **Publication-quality visuals**: ggplot2 produces attractive, polished figures suitable for scientific publications with sensible defaults [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk).
4. **Customization**: It is easier to customize legends, colors, themes, and annotations, and to combine multiple plots [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk).
5. **Data mapping**: Direct mapping of data columns to visual aesthetics (color, size, shape) is built-in and intuitive [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk).
6. **Reproducibility**: ggplot2 code is scriptable and reproducible, making it easy to regenerate or update figures [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk).

Base R plots are quick for simple, exploratory graphics, but become fiddly and time-consuming for refined, complex, or publication-ready figures [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk). 

Which of these advantages do you think is most important for your work?
