---
title: "Class 17 Q13-14"
author: "Jennifer Thai (PID: A17893762)"
format: gfm
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 
Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein.  Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding. 


```{r}
file_path <- "sample geno exp.txt"

my_table <- read.table(
      file = file_path,
      header = TRUE,  
      sep = " ",    
      skip = 0       
    )
```


```{r}
table(my_table$geno)
```

```{r}
tapply(my_table$exp, my_table$geno, median)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 
Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.  

```{r}
library(ggplot2)

ggplot(my_table, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot() +
  labs(x = "Genotype", y = "Expression") +
  geom_point()
```

G/G has a lower expression value compared to A/A and A/G at this position/locus, and the SNP does affect the expression of ORMDL3 in that a copy of the G allele will reduce the expression of this gene. This explains why G/G has the lowest expression value and why A/A has the highest expression.





